---
title: "Group3_Assessment2"
author: "06014041, 06035495, 06029131 and 02277197"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dependencies

```{r dependencies}
library(conflicted)  
library(tidyverse)  
library(lubridate)   
library(tibble)     
library(tidyr)   
library(ggplot2)  
library(data.table) 
library(visdat) 
library(naniar)     
library(zoo)        
library(GGally)       
library(corrplot)     
library(ggcorrplot)   
library(car)          
library(tseries)      
library(forecast) 
library(patchwork)
library(plotly)
library(gridExtra)
library(grid)
library(ggplotify)
library(lmtest)
library(vars)
library(urca)
library(factoextra)
library(ppcor)
library(ggExtra)
library(stats)


# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
conflict_prefer("month", "lubridate")
conflict_prefer("year", "lubridate")
conflict_prefer("wday", "lubridate")
conflict_prefer("select", "plotly")

```

# Introduction

This notebook reproduces our exploratory data analyses of UK daily energy demand (2020–2024) and its relationship to renewable energy sources, weather, and price variables.

# Load and Clean Data (Jakub)

```{r}
file_path <- "energy_demand_uk.csv"  
df <- fread(file_path)  
df$V1 <- NULL
# Preview dataset 
str(df)
glimpse(df)
summary(df)

# Convert date column to proper Date format
df$date <- as.Date(df$date, format = "%Y-%m-%d")
df$year <- year(df$date)
df$month <- month(df$date, label=TRUE)
df$day_of_week <- wday(df$date, label=TRUE)

# Convert all time series columns to numeric
numeric_cols <- c("national_demand", "wind_generation", "solar_generation", 
                  "min_temp", "max_temp", "rain_mm", "wind_speed", "average_price_daily")
df[, (numeric_cols) := lapply(.SD, as.numeric), .SDcols = numeric_cols]

sapply(df, class)  # Verify data types

#
# 1.2 Handling Missing Data
#

# Count missing values per column
colSums(is.na(df))

# Handle missing data
df[, min_temp := na.approx(min_temp, na.rm = FALSE)]
df[, max_temp := na.approx(max_temp, na.rm = FALSE)]
df[, rain_mm := rollapply(rain_mm, width = 7, FUN = mean, fill = "extend", align = "right")]
df[, wind_speed := na.approx(wind_speed, na.rm = FALSE)]


# Check if missing values arefilled
colSums(is.na(df))

#
# 1.3 Check for Duplicates & Anomalies
#

# Remove duplicate rows if they exist
df <- unique(df)

# Apply log transformation to all numeric variables
df_log <- df %>%
  mutate(across(all_of(numeric_cols), ~log1p(.)))  

# Boxplot after log transformation
df_log %>% 
  pivot_longer(cols = numeric_cols, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(outlier.colour = "red") +
  theme_minimal() +
  coord_flip() +
  ggtitle("Boxplots (Log-Transformed)")


# Identify extreme values using Z-score method
outlier_threshold <- 3  # Standard Z-score threshold
z_scores <- df[, lapply(.SD, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)), .SDcols = numeric_cols]
outliers <- df[rowSums(abs(z_scores) > outlier_threshold) > 0]

# Print num of detected outliers
print(nrow(outliers))

# Manually Winsorizing numeric columns (capping extreme values at 1st and 99th percentile)
df[, (numeric_cols) := lapply(.SD, function(x) {
  q_low <- quantile(x, 0.01, na.rm = TRUE)   # 1st percentile
  q_high <- quantile(x, 0.99, na.rm = TRUE)  # 99th percentile
  x <- pmax(q_low, pmin(x, q_high))  # Cap values within this range
  return(x)
}), .SDcols = numeric_cols]

summary(df)
str(df)

```

# Univariate EDA (Joshua)

```{r}
# Pivot the data to long format
energy_data_long <- df %>%
  pivot_longer(
    cols = c(national_demand, wind_generation, solar_generation,
             min_temp, max_temp, rain_mm, wind_speed, average_price_daily),
    names_to = "variable",
    values_to = "value"
  )

# Plot each variable as a time series, using facets
ggplot(energy_data_long, aes(x = date, y = value)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Time Series Plots of All Variables",
    x = "Date",
    y = "Value"
  ) +
  theme_bw()
```

# National Demand Univariate (Muhammad)

```{r}

ggplot(df, aes(x = date, y = national_demand)) +
  geom_line(color = "blue") +
  labs(title = "Energy Demand Over Time", x = "Date", y = "Energy Demand") + 
  theme_minimal() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y")

```

## ADF Test

```{r}
adf_test <- adf.test(df$national_demand)
print(adf_test)
```

## ACF Analysis

```{r}
acf(df$national_demand, main = "ACF of Energy Demand")
```

```{r}
pacf(df$national_demand, main = "PACF of Energy Demand")
```

## Performing seasonal differencing

```{r}
ts_data <- ts(df$national_demand, start = c(2020, 6), frequency = 365)
seasonal_diff <- diff(ts_data, lag = 365)
seasonal_diff_ts <- ts(seasonal_diff, start = c(2021, 6), frequency = 365)

start_date <- as.Date("2020-06-01")
dates_original <- seq(start_date, by = "day", length.out = length(ts_data))
dates_diff <- seq(start_date + 365, by = "day", length.out = length(seasonal_diff_ts))


z_data <- zoo(coredata(ts_data), dates_original)
z_diff <- zoo(coredata(seasonal_diff_ts), dates_diff)


par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))


plot(z_data, main = "Original Energy Demand", ylab = "Demand (MW)", 
     xaxt = "n", xlab = "")
axis.Date(1, at = seq(as.Date("2020-06-01"), as.Date("2024-06-01"), 
          by = "6 months"), format = "%b %Y", las = 1)

plot(z_diff, main = "Seasonally Differenced Energy Demand", 
     ylab = "Differenced Value", xaxt = "n", xlab = "")
axis.Date(1, at = seq(as.Date("2021-06-01"), as.Date("2024-06-01"), 
          by = "6 months"), format = "%b %Y", las = 1)

mtext("Time", side = 1, line = 3.5, outer = TRUE)
par(mfrow = c(1, 1))
```

```{r}
adf_diff <- adf.test(seasonal_diff)
print(adf_diff)
```

```{r}
acf(seasonal_diff, main = "ACF of Seasonally Differenced Energy Demand")
```

```{r}
pacf(seasonal_diff, main = "PACF of Seasonally Differenced Energy")
```

## MA Smoothing

```{r}
windows <- c(7, 14, 30, 90, 365)
colors <- c("blue", "red", "green", "purple", "orange")

par(mar = c(5, 4, 4, 10) + 0.1)  


plot(df$date, df$national_demand, type = "l", col = "gray", lwd = 1, 
     main = "Smoothing with Different Window Sizes", 
     xlab = "Date", ylab = "Energy Demand", 
     cex.main = 1.2, cex.lab = 1.1, cex.axis = 1.0)


for (i in 1:length(windows)) {
  ma <- rollmean(df$national_demand, k = windows[i], fill = NA, align = "center")
  lines(df$date, ma, col = colors[i], lwd = 2)
}


par(xpd = TRUE) 
legend("topright", inset = c(-0.45, 0), legend = paste(windows, "day MA"), 
       col = colors, lwd = 2, bty = "n", cex = 1.0)
par(xpd = FALSE)

```

## Seasonal Pattern Analysis

```{r}
# monthly_avg <- aggregate(national_demand ~ month, df, mean)
# 
# ggplot(monthly_avg, aes(x = month, y = national_demand, group = 1)) +
#   geom_line() +
#   geom_point() +
#   labs(title = "Average National Demand by Month", x = "Month", y = "Energy Demand") +
#   theme_minimal()

```

## Time Series Decomposition

```{r}
ts_decomp <- stl(ts_data, s.window = "periodic")
plot(ts_decomp)
```

## Spectral Analysis

```{r}
spectrum_result <- spectrum(df$national_demand, span = 85)
dominant_freq <- spectrum_result$freq[which.max(spectrum_result$spec)]


cat("Dominant frequency:", dominant_freq, "\n")
cat("Corresponding period (days):", 1 / dominant_freq, "\n")
```

## Yearly Comparison

```{r}
ggplot(df, aes(x = month, y = national_demand, color = factor(year), group = year)) +
  stat_summary(fun = mean, geom = "line") +
  labs(title = "Monthly Patterns by Year", x = "Month", y = "Energy Demand", color = "Year") +
  theme_minimal()
```
# National Demand Univariate (Adam)
```{r summary statistics}

# Summary statistics for the national demand column
summary(df$national_demand)

mean <- mean(df$national_demand, na.rm = TRUE)
median <- median(df$national_demand, na.rm = TRUE)
sd <- sd(df$national_demand, na.rm = TRUE)
min <- min(df$national_demand, na.rm = TRUE)
max <-max(df$national_demand, na.rm = TRUE)

paste("standard deviation", sd)


```
```{r investigation}
df_nd <- df %>% select(national_demand, date)
print(df_nd)
```
## Including Plots

```{r exploratory plots, echo=FALSE}
# Histogram of the national demand using ggplot2
distribution_plot <- ggplot(df, aes(x = national_demand)) +
  geom_histogram(binwidth = 1000, fill = "blue", color = "black") +
  labs(title = "Histogram of National Demand", x = "National Demand", y = "Frequency")

# Histogram shows a bimodal distribution of the national demand variable potentially at the winter and summer peaks, with winter being the higher peak

# Boxplot
ggplot(df, aes(y = national_demand)) +
  geom_boxplot(fill = "orange") +
  labs(title = "Boxplot of National Demand", y = "National Demand") + 
  theme_minimal()

# Time Series
Time_series_plot <- ggplot(df, aes(x = date, y = national_demand)) +
  geom_line(color="blue") +
  #geom_smooth(method = "glm") +
  labs(title = "National Energy Demand Time Series (2020-2024)", x = "Date", y = "National Energy Demand (MW)") +
  theme_minimal()

# Clear cyclical element to energy demand caused by a seasonal effect of higher demand of energy in winter

```

```{r Time Series Analysis}
# A time series is stationary if it fulfills the following conditions:
# - The expected value of Y is constant over time
# - the variance of y is finite
# - the covariance of yt and y(t + h) depends only on the value of h = 0,1,2,.. for all t
# 
# Looks broadly stationary but let's inspect further by performing an Augmented Dickey Fuller test
# Perform the Augmented Dickey-Fuller test to check for stationary
adf_test <- adf.test(df$national_demand, alternative = "stationary")
adf_test

# reject the null hypothesis of the ADF test that the time series is stationary so we can assume that the time series is stationary at a 5% significance level.

acf(df$national_demand, main = "Autocorrelation Plot of National Demand")

```

```{r Seasonality Analysis}

df_spec <- df %>% drop_na()
energy_demand <- df$national_demand

old_par <- par(mfrow = c(2, 2))

plot(energy_demand) # data in time domain

spectrum_result <- spectrum(energy_demand, plot = FALSE)

spectrum(energy_demand, spans = 20)

common_frequency  <- spectrum_result$freq[which.max(spectrum_result$spec)]

common_frequency

paste("Spectral Analysis implies a periodicity of: ", 1/common_frequency)
```
# Initial Multivariate and Causal Analysis (Jacob)
```{r}
# 2 Initial Multivariate exploration
df
# Select only numeric variables for correlation and multivariate analysis
numeric_vars <- df %>% select(-date, -year, -month, -day_of_week) # Remove non-numeric
numeric_vars
# 
# Pairwise Correlation Analysis
# 
numeric_vars
# Compute correlation matrix ## UNUSED
cor_matrix <- cor(numeric_vars, use="complete.obs", method="pearson") # Note to marker: spearman has been used for bersion included in report to allow for potential non-linear relationships. This matrix was unused here

# Plot the UNUSED correlation matrix
ggcorrplot(cor_matrix, 
           method = "circle", 
           type = "upper", 
           lab = TRUE, 
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix of Energy Demand Dataset")

##############################################################################################

# Custom correlation panel (upper panels)
custom_cor <- function(data, mapping, ...) {
  x <- eval_data_col(data, mapping$x)
  y <- eval_data_col(data, mapping$y)
  corr <- cor(x, y, use = "complete.obs")
  corr_text <- sprintf("%.2f", corr)
  
  df <- data.frame(x = 1, y = 1, corr = corr)
  
  ggplot(df, aes(x, y, fill = corr)) +
    geom_tile(color = "white") +
    geom_text(aes(label = corr_text), size = 5) +
    scale_fill_gradient2(
      low = "blue",     # Negative correlation
      mid = "white",    # Zero correlation
      high = "red",     # Positive correlation
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_void() +
    theme(legend.position = "right")
}


my_column_labels <- c(
  "National\nDemand",
  "Wind\nGen",
  "Solar\nGen",
  "Min\nTemp",
  "Max\nTemp",
  "Rain\nmm",
  "Wind\nSpeed",
  "Avg\nPrice\nDaily"
)

options(scipen = 999)
numeric_vars
# Create the ggpairs plot
p <- ggpairs(
  numeric_vars,
  columnLabels = my_column_labels,  
  lower = list(
    continuous = wrap("points", alpha = 0.3, size = 0.8, color = "gray40")
  ),
  diag = list(
    continuous = wrap("densityDiag", fill = "gray40", color = "gray40", alpha = 0.3)
  ),
  upper = list(
    continuous = custom_cor
  ),
  title = "Pairwise Scatterplot with Correlation Matrix"
)

# Adjust theme to fix label overlap & give more space
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
  axis.text.y = element_text(size = 8),
  strip.text  = element_text(size = 8),
  plot.margin = margin(10, 10, 10, 10)
)


print(p)



#PCA
# Standardize data 
df_scaled <- scale(numeric_vars)

# Compute PCA
pca_model <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

#Visualize PCA Biplot
fviz_pca_biplot(pca_model, label = "var", col.var = "red", col.ind = "blue",
                title = "PCA Biplot of Energy Demand Data")

#Scree Plot (Variance Explained)
fviz_eig(pca_model, addlabels = TRUE, title = "Scree Plot of PCA")

# PCA Contributions
fviz_pca_var(pca_model, col.var = "contrib",
             gradient.cols = c("blue", "red"),
             title = "Variable Contributions to PCA")

# 
# 4. Adjusted Multicollinearity Analysis (VIF)
# 

# Drop max_temp (because of 0.9 correlation with min_temp)
#vif_df <- numeric_vars %>% select(-c(max_temp)) 
vif_df <- numeric_vars 
# Compute Variance Inflation Factor (VIF)
vif_results <- vif(lm(national_demand ~ ., data = vif_df))

# Print VIF results
print(vif_results)

#Drop max_temp (because of 0.9 correlation with min_temp)
vif_df <- numeric_vars %>% select(-c(max_temp)) 

# # Set up a grid layout: 3 rows and 2 columns
# par(mfrow = c(3, 2))
# 
# # Cross-Correlation: Energy Demand vs. Wind Generation
# ccf(energy_ts, wind_ts, lag.max = 30, main = "CCF: Energy Demand vs. Wind Generation")
# 
# # Cross-Correlation: Energy Demand vs. Solar Generation
# ccf(energy_ts, solar_ts, lag.max = 30, main = "CCF: Energy Demand vs. Solar Generation")
# 
# # Cross-Correlation: Energy Demand vs. Temperature
# ccf(energy_ts, temp_ts, lag.max = 30, main = "CCF: Energy Demand vs. Temperature")
# 
# # Cross-Correlation: Energy Demand vs. Rainfall
# ccf(energy_ts, rain_ts, lag.max = 30, main = "CCF: Energy Demand vs. Rainfall")
# 
# # Cross-Correlation: Energy Demand vs. Wind Speed
# ccf(energy_ts, wind_speed_ts, lag.max = 30, main = "CCF: Energy Demand vs. Wind Speed")
# 
# # Cross-Correlation: Energy Demand vs. Electricity Price
# ccf(energy_ts, price_ts, lag.max = 30, main = "CCF: Energy Demand vs. Electricity Price")
# 
# 
# # par(mfrow = c(1, 1))
# #
# # ccf(wind_ts, temp_ts, lag.max = 30, main = "CCF: Wind energy Generation vs. Temperature")
# 
# # =====================================
# # 5.2 GRANGER CAUSALITY ANALYSIS
# # 
# optimal_lag <- VARselect(cbind(energy_ts, solar_ts), lag.max = 150, type = "const")
# print(optimal_lag)
# 
# # Granger Test: Does Temperature Granger-Cause Energy Demand?
# grangertest(energy_ts ~ temp_ts, order = 9)
# 
# # Granger Test: Does Temperature Granger-Cause Energy Demand?
# grangertest(solar_ts ~ temp_ts, order = 9)
# 
# # Granger Test: Does Temperature Granger-Cause Energy Demand?
# grangertest(temp_ts ~ solar_ts, order = 9)
# 
# c# Granger Test: Does Wind Generation Granger-Cause Energy Demand?
# grangertest(energy_ts ~ wind_ts, order = 9)
# 
# # Granger Test: Does Solar Generation Granger-Cause Energy Demand?
# grangertest(energy_ts ~ solar_ts, order = 5)
# 
# # Granger Test: Does Energy Demand Granger-Cause Electricity Price?
# grangertest(price_ts ~ energy_ts, order = 5)
# 
# # Granger Test: Does Electricity Price Granger-Cause Energy Demand?
# grangertest(energy_ts ~ price_ts, order = 5)
# 
# # Granger Test: Rainfall Granger-Cause Energy Demand?
# grangertest(wind_ts ~ energy_ts, order = 5)
# 
# # Granger Test: Rainfall Granger-Cause Energy Demand?
# grangertest(wind_ts ~ solar_ts, order = 5)
# 
# # =====================================
# # 5.3 IDENTIFY OPTIMAL LAG STRUCTURES
# # =====================================
# 
# # **Impulse Response Function (IRF)**
# irf_model <- irf(var_model, impulse = "solar", response = "energy", n.ahead = 30)
# plot(irf_model)
# 
# # **Plot 3: Interactive Multivariate Plot**
# p3 <- ggplot(df, aes(x = date)) +
#   geom_line(aes(y = national_demand, color = "Energy Demand"), size = 1) +
#   geom_line(aes(y = wind_generation, color = "Wind Generation"), size = 1) +
#   geom_line(aes(y = solar_generation, color = "Solar Generation"), size = 1) +
#   labs(title = "Interactive Multivariate Time Series Plot", x = "Date", y = "Value") +
#   theme_minimal()
# 
# interactive_p3 <- ggplotly(p3)  # Convert to interactive plot
# interactive_p3
# 
# # Vector Autoregression (VAR) to Determine Optimal Lags
# data_var <- data.frame(energy_ts, solar_ts)
# 
# # Find Optimal Lag Length for VAR Model
# lag_selection <- VARselect(data_var, lag.max = 365, type = "const")
# 
# # Print Recommended Lag
# print(lag_selection$selection)
# 
# packages <- c("tidyverse", "tseries", "urca", "vars", "forecast")
# # install.packages(setdiff(packages, installed.packages()[,"Package"]), dependencies = TRUE)
# 
# # =====================================
# # 7.1 VECTOR AUTOREGRESSION (VAR)
# # =====================================
# 
# # **Check for Stationarity using Augmented Dickey-Fuller (ADF) Test**
# adf_energy <- adf.test(energy_ts)
# adf_wind   <- adf.test(wind_ts)
# adf_solar  <- adf.test(solar_ts)
# 
# 
# # Print results
# print(adf_energy)
# print(adf_wind)
# print(adf_solar)
# 
# 
# # **If non-stationary, take first differences**
# diff_solar  <- diff(solar_ts)
# 
# adf_solar  <- adf.test(diff_solar)
# print(adf_solar)
# 
# 
# # Combine into a data frame for VAR modeling
# var_data <- data.frame(
#   energy = energy_ts[-1],  # Remove first row
#   wind   = wind_ts[-1],    # Remove first row
#   solar  = diff_solar      # Already differenced, so it matches
# )
# 
# # **Fit a VAR Model**
# var_model <- VAR(var_data, p = 29, type = "const")  # Using lag = 5 (adjust if needed)
# summary(var_model)
# 
# 
# 
# # =====================================
# # 7.2 VECTOR ERROR CORRECTION MODEL (VECM)
# # =====================================
# 
# # **Test for Cointegration using Johansen Test**
# johansen_test <- ca.jo(var_data, type = "trace", ecdet = "const", K = 5)
# summary(johansen_test)
# 
# # **If Cointegration Exists, Fit a VECM Model**
# if (johansen_test@teststat[2] > johansen_test@cval[2,2]) {  # Checking significance of the test
#   vecm_model <- cajorls(johansen_test, r = 1)  # Rank = 1 means at least one cointegrating relationship
#   print(summary(vecm_model))
# } else {
#   print("No significant cointegration found. Proceeding without VECM.")
# }
```
# Further Multivariate EDA Visualisations (Josh)
# Note to marker: this section can be skipped as none of these plots made it to the final report or presentation.The key multivariate
# code plots can be found in "Final Multivariate Plots (Josh)
```{r}
#--------------------------------
# Split by Winter and Summer Months
#--------------------------------
# Create monthly column
library(lubridate)
library(dplyr)

# Create categorical columns for each month
df_monthly <- df %>%
  mutate(Month = factor(month(date), levels = 1:12, 
                        labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")))

# Boxplot for monthly demand
ggplot(df_monthly, aes(x = Month, y = national_demand)) +
  geom_boxplot(outlier.color = "black") +
  labs(
    title = "Monthly Boxplots of National Demand",
    x = "Month",
    y = "Demand (MW)"
  ) +
  theme_minimal()

# Boxplot for monthly solar generation
ggplot(df_monthly, aes(x = Month, y = solar_generation)) +
  geom_boxplot(outlier.color = "black") +
  labs(
    title = "Monthly Boxplots of Solar Generation",
    x = "Month",
    y = "Solar Generation (MW)"
  ) +
  theme_minimal()

# Create categorical columns for each season
df_seasonal <- df %>%
  mutate(Season = case_when(
    month(date) %in% 3:5 ~ "Spr",
    month(date) %in% 6:8 ~ "Sum",
    month(date) %in% 9:11 ~ "Aut",
    TRUE ~ "Win"
  ))

# Boxplot for seasonal demand
ggplot(df_seasonal, aes(x = Season, y = national_demand)) +
  geom_boxplot(outlier.color = "black") +
  labs(
    title = "Seasonal Boxplots of National Demand",
    x = "Season",
    y = "Demand (MW)"
  ) +
  theme_minimal()

# Boxplot for seasonal solar generation
ggplot(df_seasonal, aes(x = Season, y = solar_generation)) +
  geom_boxplot(outlier.color = "black") +
  labs(
    title = "Seasonal Boxplots of Solar Generation",
    x = "Season",
    y = "Solar Generation (MW)"
  ) +
  theme_minimal()
# --------------------------------
# Stationarity Checks
# --------------------------------
# ADF Tests
library(tseries)

# For national_demand

help(adf.test)

adf_result_demand <- adf.test(df$national_demand, alternative = "stationary")
print(adf_result_demand) # reject this has a unit root, likely stationary

# For solar_generation
adf_result_solar <- adf.test(df$solar_generation, alternative = "stationary")
print(adf_result_solar) # fail to reject this has a unit root, likely non-stationary

# --------------------------------
# Seasonal Decomposition
# --------------------------------
df
# Convert to ts object with daily frequency ~ 365
demand_ts_national <- ts(df$national_demand, frequency = 365, start = c(2020, 1))

# Perform STL decomposition (handles daily data if we treat frequency=365)
decomp_result <- stl(demand_ts_national, s.window = "periodic")  
plot(decomp_result, main = "STL Decomposition of National Demand")
help(stl)

# Stationarity Checks after decomposition ADF Tests
adf_result_demand_std <- adf.test(df$national_demand, alternative = "stationary")
print(adf_result_demand_std) # reject this has a unit root, likely stationary

# Convert for solar generation
solar_ts <- ts(df$solar_generation, frequency = 365, start = c(2020, 1))

# Perform STL decomposition (handles daily data if we treat frequency=365)
decomp_result_solar <- stl(solar_ts, s.window = "periodic")

# Convert for national_demand
adf_result_demand <- adf.test(df$national_demand, alternative = "stationary")
print(adf_result_demand) # reject this has a unit root, likely stationary

# Convert for average_price
price_ts <- ts(df$average_price_daily, frequency = 365, start = c(2020, 1))

# Perform STL decomposition (handles daily data if we treat frequency=365)
decomp_result_price <- stl(price_ts, s.window = "periodic")

#================================================================================================
# Final Multivariate Visualisations
#================================================================================================
str(df)
# --------------------------------
# Setup
# --------------------------------
# Create "Month", "Season", and "Period" columns
# prefer lubridate::month

df_ext <- df %>%
  mutate(
    Month = factor(month(date), 
                   levels = 1:12, 
                   labels = c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")),
    Season = case_when(
      month(date) %in% 3:5  ~ "Spr",
      month(date) %in% 6:8  ~ "Sum",
      month(date) %in% 9:11 ~ "Aut",
      TRUE ~ "Win"
    ),
    Period = case_when(
      month(date) %in% c(10,11,12,1,2,3) ~ "Oct–Mar",
      TRUE ~ "Apr–Sep"
    ),
    Period_Solar = case_when(
      month(date) %in% c(11,12,1,2) ~ "Nov-Feb",
      TRUE ~ "Mar-Oct"
    ),
    
    log_solar_gen = log1p(
      solar_generation),
    
    log_wind_gen = log1p(
      wind_generation),
    
    log_national_demand = log1p(
      national_demand)
  )
# All are now factors
str(df_ext)

# # Add log transformation of solar generation
# df_ext <- df_ext %>%
#   mutate(log_solar_gen = log(solar_generation))

# Colour blind friendly pallette
cbbPalette <- c("#000000","#E69F00","#56B4E9","#009E73",
                "#F0E442","#0072B2","#D55E00","#CC79A7")

df_ext
# --------------------------------
# Uni-variate Visualisations
# --------------------------------
# Boxplot for national_demand



# 1. Histograms + Kernel Density
  plot_hist_density <- function(data, var_name, fill_color = "viridis") {
    # Basic histogram + density overlay
    ggplot(data, aes(x = .data[[var_name]])) +
      geom_density(alpha = 0.3, fill = "gray") +
      scale_fill_viridis_c(option = "magma") +  # color-blind friendly
      labs(title = paste("Distribution of ", var_name),
           x = var_name, y = "Density") +
      theme_minimal()
  }
  
  # Plot for for each variable:
  plot_hist_density(df_ext, "national_demand")
  plot_hist_density(df_ext, "solar_generation")
  plot_hist_density(df_ext, "min_temp")
  
  
  # Plot log solar generation
  plot_hist_density(df_ext, "log_solar_gen")

  df_ext
  # Combined distribution plot
  # Prefer plotly::select
  
  df_forplot <- df_ext %>%
    select(wind_generation, solar_generation)
  
  # Pivot longer to “variable” and “value” columns:
  df_melt <- df_forplot %>%
    pivot_longer(cols = everything(), 
                 names_to = "Variable", 
                 values_to = "Value")

  # 2 univariate solar and wind
  ggplot(df_melt, aes(x = Value, fill = Variable)) +
    geom_density(alpha = 0.3) +
    scale_fill_manual(values = c("#56B4E9", "#D55E00")) +
    labs(title = "Distribution of Wind and Solar Generation)#",
         x = "Value", y = "Density") +
    theme_minimal()
  
  # 2 univariate demand and wind
  df_forplot2 <- df_ext %>%
    select(wind_generation, national_demand)
  
  df_melt_demand_wind <- df_forplot2 %>%
    pivot_longer(cols = everything(), 
                 names_to = "Variable", 
                 values_to = "Value")
  
  ggplot(df_melt_demand_wind, aes(x = Value, fill = Variable)) +
    geom_density(alpha = 0.3) +
    scale_fill_manual(values = c("#56B4E9", "#D55E00")) +
    labs(title = "Distribution of National Demand and Wind") +
    theme_minimal()

  
  # 3. Violin Plots (univariate)
  
  plot_violin <- function(data, var_name) {
    ggplot(data, aes(x = "", y = .data[[var_name]])) +
      geom_violin(color = "black") +
      geom_boxplot(width = 0.2, outlier.color = "black", alpha = 0.4) +
      labs(title = paste("Violin Plot of", var_name),
           x = NULL, y = var_name) +
      theme_minimal()
  }
  
  plot_violin(df_ext, "national_demand")
  plot_violin(df_ext, "solar_generation")
  plot_violin(df_ext, "min_temp")
  plot_violin(df_ext, "log_solar_gen")

# --------------------------------
# Univariate Plots by Month/Season/Period
# --------------------------------

# 1. Violin by Month

# # National Demand
ggplot(df_ext, aes(x = Month, y = national_demand)) +
  geom_violin(color = "black", alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: National Demand by Month",
       x = "Month", y = "MW")
  
# Solar Generation
ggplot(df_ext, aes(x = Month, y = solar_generation)) +
  geom_violin(color = "black", alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Solar Generation by Month",
       x = "Month", y = "MW")

# Check for 0 values for solar generaiton
df_ext %>%
  filter(solar_generation == 0) %>%
  select(date, national_demand, solar_generation) %>%
  arrange(date) %>%
  head(20)  # Check first 20 occurrences

# Return month labels that had 0 solar generation
df_ext %>%
  filter(solar_generation == 0) %>%
  pull(Month) %>%
  unique()

df_ext %>%
  filter(solar_generation == 0) %>%
  count(Month, name = "Zero_Generation_Days")

df_ext %>%
  filter(solar_generation == 0) %>%
  mutate(Year = lubridate::year(date)) %>%  # Extract Year from date
  count(Year, Month, name = "Zero_Generation_Days") %>%
  arrange(Year, Month)  # Order results by Year and Month

# Every year from November to February had 28-31 zero-generation fays

# Log Solar Generation
ggplot(df_ext, aes(x = Month, y = log_solar_gen)) +
  geom_violin(color = "black", alpha = 0.7) +
  labs(title = "Violin: Log(Solar Generation) by Month",
       x = "Month", y = "MW") +
  theme_minimal()

# Solar Generation Boxplot
ggplot(df_ext, aes(x = Month, y = solar_generation)) +
  geom_boxplot(outlier.color = "black") +
  labs(title = "Boxplot of Solar Generation by Month",
       x = "Month", y = "Solar Generation") +
  theme_minimal()

# Log Solar Generation Boxplot
ggplot(df_ext, aes(x = Month, y = log_solar_gen)) +
  geom_boxplot(outlier.color = "black") +
  labs(title = "Boxplot of Log(Solar Generation) by Month",
       x = "Month", y = "Log(Solar Generation)") +
  theme_minimal()

# min_temp
ggplot(df_ext, aes(x = Month, y = min_temp)) +
  geom_violin(color = "black", alpha = 0.7) +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Min Temperature by Month",
       x = "Month", y = "Min Temperature (°C)") +
  theme_minimal()

# 2. Violin by Season

# # National Demand
# ggplot(df_ext, aes(x = Season, y = national_demand)) +
#   geom_violin(aes(fill = Season), color = "black", alpha = 0.7) +
#   scale_fill_viridis_d(option = "plasma") +
#   geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
#   labs(title = "Violin: National Demand by Season",
#        x = "Season", y = "MW") +
#   theme_minimal()
# 
# # Solar Generation
# ggplot(df_ext, aes(x = Season, y = solar_generation)) +
#   geom_violin(aes(fill = Season), color = "black", alpha = 0.7) +
#   scale_fill_viridis_d(option = "plasma") +
#   geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
#   labs(title = "Violin: Solar Generation by Season",
#        x = "Season", y = "MW") +
#   theme_minimal()

# 3. Violin by Period (Oct–Mar vs. Apr–Sep)

# National Demand
ggplot(df_ext, aes(x = Period, y = national_demand)) +
  geom_violin() +
  # scale_fill_viridis_d(option = "D") +
  geom_boxplot(width = 0.2, alpha = .7) +
  labs(x = "Period", y = "National Demand (MW)")

# National Demand (no period)
ggplot(df_ext, aes(x = "", y = national_demand)) +
  geom_violin() +
  # scale_fill_viridis_d(option = "D") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  labs(x = NULL, y = "National Demand (MW)")

help(scale_fill_viridis_d)
help(aes)

# Solar Generation
ggplot(df_ext, aes(x = Period, y = solar_generation)) +
  geom_violin() +
  scale_fill_viridis_d(option = "plasma") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Solar Generation by Period",
       x = "Period", y = "Solar Generation (MW)") +
  theme_minimal()

# Wind Generation
ggplot(df_ext, aes(x = Period, y = wind_generation)) +
  geom_violin() +
  scale_fill_viridis_d(option = "plasma") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Wind Generation by Period",
       x = "Period", y = "Wind Generation (MW)") +
  theme_minimal()


# Temperature
ggplot(df_ext, aes(x = Period, y = min_temp)) +
  geom_violin() +
  # scale_fill_viridis_d(option = "plasma") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Min Temperature by Period",
       x = "Period", y = "Min Temperature (°C)") +
  theme_minimal()

# Log solar generation
ggplot(df_ext, aes(x = Period, y = log_solar_gen)) +
  geom_violin() +
  scale_fill_viridis_d(option = "plasma") +
  geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.3) +
  labs(title = "Violin: Log(Solar Generation) by Period",
       x = "Period", y = "Log(Solar Generation)") +
  theme_minimal()


# Demand Solar and Wind
# Tidy format
df_tidy <- df_ext %>%
  select(Period, national_demand, solar_generation, wind_generation,) %>%
  pivot_longer(
    cols = c(national_demand, solar_generation, wind_generation),
    names_to = "variable",
    values_to = "value"
  )

# Box plot combined by period
ggplot(df_tidy, aes(x = Period, y = value)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y")

# Violin by period
ggplot(df_tidy, aes(x = Period, y = value)) +
  geom_violin() +
  facet_wrap(~ variable, scales = "free_y")

# Box combined with violin addition
# 1) Tidy data with 'Period' and 'variable', 'value'
df_tidy <- df_ext %>%
  select(Period, national_demand, solar_generation, wind_generation) %>%
  pivot_longer(
    cols = c(national_demand, solar_generation, wind_generation),
    names_to = "variable",
    values_to = "value"
  )

# 2) Plot a single violin ignoring Period, plus separate boxplots by Period
ggplot() +
  facet_wrap(~ variable, scales = "free_y") +
  
  # (a) Single violin for entire distribution, ignoring Period
  geom_violin(
    data = df_tidy,
    aes(x = "All", y = value),
    fill = "gray80", alpha = 0.3, color = "black", linewidth = 0.5
  ) +
  
  # (b) Boxplots split by Period
  geom_boxplot(
    data = df_tidy,
    aes(x = Period, y = value),
    width = 0.4, outlier.shape = 16
  ) +
  
  # 3) Force the x scale to have three discrete slots: "All", "Apr–Sep", "Oct–Mar"
  scale_x_discrete(
    name = NULL,
    limits = c("All", "Apr–Sep", "Oct–Mar") 
  ) +
  
  # labs(
  #   title = "Violin of Entire Distribution vs. Boxplots by Period"
  # ) +
  theme_minimal()


  
# Box plot combined overall
ggplot(df_tidy, aes(x = "", y = value)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y")

help(facet_wrap)
# Log 

ggplot(df_tidy, aes(x = variable, y = value, fill = Period)) +
  geom_violin(position = position_dodge(width=0.8)) +
  scale_y_log10() +
  labs(title = "Violin: Demand vs. Solar vs. Wind on Log Scale")

# Tidy with logs
df_tidy_dem_sol <- df_ext %>%
  select(Period, national_demand, solar_generation) %>%
  pivot_longer(
    cols = c(national_demand, solar_generation),
    names_to = "variable",
    values_to = "value"
  )
#REMOVED
# df_tidy_all_log <- df_ext %>%
#   select(log_national_demand, log_solar_gen, log_wind_gen) %>%
#   pivot_longer(
#     cols = c(log_national_demand, log_solar_gen, log_wind_gen),
#     names_to = "variable",
#     values_to = "value"
#   )

# Box plot combbined by period Demand vs Solar
ggplot(df_tidy_dem_sol, aes(x = Period, y = value)) +
  geom_boxplot() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Boxplots by Variable and Period in MW")

# REMOVED
# # Demand Log_Solar and Wind 
# # Tidy format just solar log
# df_tidy_log_solar <- df_ext %>%
#   mutate(
#     variable = factor(variable,
#                       levels = c("national_demand", "log_solar_gen", "wind_generation")
#     )
#   )
# 
# 
# # Violins
# ## Just solar log
# ggplot(df_tidy_log_solar, aes(x = Period, y = value)) +
#   geom_violin() +
#   facet_wrap(~ variable, scales = "free_y") +
#   labs(title = "national_demand (MW) vs. log_solar_gen (log(MW)) vs. wind_generation (MW)")



## Just solar log
ggplot(df_tidy, aes(x = variable, y = value, fill = Period)) +
  geom_violin(position = position_dodge(width=0.8)) +
  scale_y_log10() +
  labs(title = "Violin: Demand vs. Solar vs. Wind on Log Scale")


## Boxplots

# REMOVED
# ggplot(df_tidy_log_solar, aes(x = Period, y = value)) +
#   geom_boxplot() +
#   facet_wrap(~ variable, scales = "free_y") +
#   labs(title = "Boxplots: log_solar_gen (log(MW)) vs. Demand (MW) vs. Wind (MW) by Period")

# # All log REMOVED
# ggplot(df_tidy_all_log, aes(x = Period, y = value)) +
#   geom_boxplot() +
#   facet_wrap(~ variable, scales = "free_y") +
#   labs(title = "Boxplots: log_solar_gen (log(MW)) vs. Demand (MW) vs. Wind (MW) by Period")
# 
# df_tidy_all_log
# df_tidy_log_solar

#  Solar and Wind
# Tidy
df_tidy_no_demand <- df_ext %>%
  select(Period, solar_generation, wind_generation,) %>%
  pivot_longer(
    cols = c(solar_generation, wind_generation),
    names_to = "variable",
    values_to = "value"
  )

ggplot(df_tidy_no_demand, aes(x = Period, y = value)) +
  geom_violin() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Violin Plots by Variable and Period")

ggplot(df_tidy_no_demand, aes(x = Period, y = value)) +
  geom_violin() +
  facet_wrap(~ variable, scales = "free_y") +
  labs(title = "Violin Plots by Variable and Period")
help(facet_wrap)
# Log 

ggplot(df_tidy_no_demand, aes(x = variable, y = value, fill = Period)) +
  geom_violin(position = position_dodge(width=0.8)) +
  scale_y_log10() +
  labs(title = "Violin: Solar vs. Wind on Log Scale")

# Scale adjustments
df_scaled <- df_ext %>%
  mutate(
    solar_scaled = solar_generation * 30,
    wind_scaled  = wind_generation * 10
  )

df_tidy_scaled <- df_scaled %>%
  select(Period, national_demand, solar_scaled, wind_scaled) %>%
  pivot_longer(
    cols = c("national_demand", "solar_scaled", "wind_scaled"),
    names_to = "variable",
    values_to = "value"
  )


# 3.1 Violin by Period (Mar-Oct vs. Nov-Feb) - Solar Generation

# --------------------------------
# Bivariate plots
# --------------------------------
# 1. General
# install.packages("ggExtra")
library(ggExtra)

# Colour blind friendly pallette
cbbPalette <- c("#000000","#E69F00","#56B4E9","#009E73",
                "#F0E442","#0072B2","#D55E00","#CC79A7")

# Demand vs Solar Gen
p_scatter <- ggplot(df_ext, aes(x = national_demand, y = solar_generation)) +
  geom_point(alpha = 0.3) +
  labs(title = "Demand vs. Solar Generation",
       x = "National Demand (MW)",
       y = "Solar Generation (MW)") +
  theme_minimal()

# Add violin marginal distributions
ggExtra::ggMarginal(p_scatter, type = "violin", fill = "#000000", alpha = .1)
ggExtra::ggMarginal(p_scatter, type = "violin", fill = "#000000", alpha = .2)

help(ggMarginal)

# Demand vs Wind
p_scatter_wind <- ggplot(df_ext, aes(x = national_demand, y = wind_generation)) +
  geom_point(alpha = 0.3) +
  labs(title = "Demand vs. Wind Generation",
       x = "National Demand (MW)",
       y = "Wind Generation (MW)") +
  theme_minimal()

# Add violin marginal distributions
ggExtra::ggMarginal(p_scatter_wind, type = "violin", fill = "#56B4E9")

# Demand vs Min Temp
p_scatter_temp <- ggplot(df_ext, aes(x = national_demand, y = min_temp)) +
  geom_point(alpha = 0.3) +
  labs(title = "Demand vs. Min Temperature",
       x = "National Demand (MW)",
       y = "Min Temperature (°C)") +
  theme_minimal()

# Add violin marginal distributions
ggExtra::ggMarginal(p_scatter_temp, type = "violin", fill = "#56B4E9")


# # 2. Separate scatter for each season
# p_scatter_season <- ggplot(df_ext, aes(x = national_demand, y = solar_generation)) +
#   geom_point(alpha = 0.4) +
#   facet_wrap(~ Season) +
#   labs(title = "Demand vs. Solar Generation by Season",
#        x = "National Demand",
#        y = "Solar Generation") +
#   theme_minimal()
# 
# p_scatter_season
# 
# # With Colours
# p_scatter_color <- ggplot(df_ext, aes(x = national_demand, y = solar_generation, color = Season)) +
#   geom_point(alpha = 0.6) +
#   scale_color_viridis_d(option = "plasma") +
#   labs(title = "Demand vs. Solar Generation (Coloured by Season)",
#        x = "National Demand", y = "Solar Generation") +
#   theme_minimal()
# 
# # Add violin or box marginal:
# ggExtra::ggMarginal(p_scatter_color, type = "violin", groupColour = TRUE, groupFill = TRUE)

# 3. Bivariate Plots by Period

# One panel per period
ggplot(df_ext, aes(x = national_demand, y = solar_generation)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~ Period) +
  labs(title = "Demand vs. Solar Generation",
       x = "National Demand (MW)", y = "Solar Generation (MW)") +
  theme_minimal()

# combined scatter with color for periods:
p_scatter_period <- ggplot(df_ext, aes(x = national_demand, y = solar_generation, color = Period)) +
  geom_point(alpha = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Demand vs. Solar Generation",
       x = "National Demand (MW)", y = "Solar Generation (MW)") +
  theme_minimal()

ggMarginal(p_scatter_period, type = "violin", groupColour = TRUE, groupFill = TRUE)

# combined scatter with color for periods (no marginal):
p_scatter_period <- ggplot(df_ext, aes(x = national_demand, y = solar_generation, color = Period)) +
  geom_point(alpha = 0.4) +
  scale_fill_brewer(palette = "") +
  labs(title = "Demand vs. Solar Generation",
       x = "National Demand (MW)", y = "Solar Generation (MW)") +
  theme_minimal()

p_scatter_period

ggExtra::ggMarginal(p_scatter_period, type = "violin")
help(ggMarginal)


# Log solar generation equivalent
p_scatter_log <- ggplot(df_ext, aes(x = national_demand, y = log_solar_gen, color = Period)) +
  geom_point(alpha = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Demand vs. Log(Solar Generation)",
       x = "National Demand (MW)", y = "Log(Solar Generation)") +
  theme_minimal()

ggMarginal(p_scatter_log, type = "violin", groupColour = TRUE, groupFill = TRUE)
ggMarginal(p_scatter_log, type = "boxplot", groupColour = TRUE, groupFill = TRUE)
ggMarginal(p_scatter_log, type = "densigram", groupColour = TRUE, groupFill = TRUE)
ggMarginal(p_scatter_log, type = "densigram", groupColour = TRUE, groupFill = TRUE)


# Demand vs Temperature
ggplot(df_ext, aes(x = national_demand, y = min_temp)) +
  geom_point(alpha = 0.4) +
  facet_wrap(~ Period) +
  labs(title = "Demand vs. Min Temperature",
       x = "National Demand (MW)", y = "Min Temperature (°C)") +
  theme_minimal()

# combined scatter with color for periods:
p_scatter_temp_period <- ggplot(df_ext, aes(x = national_demand, y = min_temp, color = Period)) +
  geom_point(alpha = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Demand vs. Min Temperature",
       x = "National Demand (MW)", y = "Min Temperature (°C)") +
  theme_minimal()

ggMarginal(p_scatter_temp_period, type = "violin", groupColour = TRUE, groupFill = TRUE)

# --------------------------------
# 4. Pairwise Scatterplots
# --------------------------------

# Colour blind friendly pallette
cbbPalette <- c("#000000","#E69F00","#56B4E9","#009E73",
                "#F0E442","#0072B2","#D55E00","#CC79A7")

df_ext

# define a smaller data set
df_gpairs <- df_ext %>%
  dplyr::select(national_demand,#1  
                solar_generation,#2
                min_temp,#3
                log_solar_gen,#4
                Period,#5
                wind_generation,#6
                Month, #7
                Season) #8

# # 1. ggpairs coloured by seasob
# ggpairs(df_gpairs,
#         columns = 1:3,   # numeric columns (demand, solar, min_temp)
#         aes(color = Season, alpha = 0.5)) +
#   scale_color_viridis_d(option = "plasma") +
#   labs(title = "Pairwise Scatterplots: Demand, Solar, Min Temp") +
#   theme_minimal()
df_gpairs

# 2. ggpairs coloured by period
ggpairs(df_gpairs,
        columns = 1:3,
        upper = list(continuous = wrap("cor", method = "spearman")), # spearman allows for non-linear relationships (warnings are ok, p values approx)
        aes(color = Period, alpha = 0.5)) +
  labs(title = "Pairwise Scatterplots: Demand, Solar Generation , Temperature") +
  theme_minimal()

# 3. ggpairs with log(solar_generation) by period
str(df_gpairs)              # Check structure
colSums(is.na(df_gpairs))   # Ensure no column is all NA


# demand solar wind and temp
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "wind_generation", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships

)

# demand solar wind and temp and season
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "min_temp", "Season"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = "cor")
)

# demand solar temp
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = "cor")
)

# wind temp and solar Demand
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "wind_generation", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships
)

# wind temp and solar Season
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "wind_generation", "min_temp", "Season"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships
)

# Wind solar demand season
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "wind_generation", "Season"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships
)
help(ggpairs)
# solar and season
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "Season"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = "cor")
)

help(ggpairs)

help(scale_color_viridis_d)

# Colour blind friendly pallette
cbbPalette <- c("#000000","#E69F00","#56B4E9","#009E73",
                "#F0E442","#0072B2","#D55E00","#CC79A7")

# ggpairs demand vs temperature
ggpairs(df_gpairs,
        columns = c(1, 3),   # demand, min_temp
        aes(color = Period, alpha = 0.5)) +
  labs(title = "Pairwise Scatterplots: Demand, Temperature") +
  theme_minimal()

# ggpairs demand vs wind
ggpairs(df_gpairs,
        columns = c(1, 6),   # demand, wind
        aes(color = Period, alpha = 0.5)) +
  labs(title = "Pairwise Scatterplots: Demand, Wind") +
  theme_minimal()

help(ggpairs)

df_gpairs

summary(df_gpairs$solar_generation)

# --------------------------------
# 4. Pairwise Scatterplots with Histograms
# --------------------------------
df_ext

# define a smaller data set
df_gpairs <- df_ext %>%
  dplyr::select(national_demand, solar_generation, min_temp, log_solar_gen, Period)

# # 1. ggpairs coloured by seasob
# ggpairs(df_gpairs,
#         columns = 1:3,   # numeric columns (demand, solar, min_temp)
#         aes(color = Season, alpha = 0.5)) +
#   scale_color_viridis_d(option = "plasma") +
#   labs(title = "Pairwise Scatterplots: Demand, Solar, Min Temp") +
#   theme_minimal()

# 2. ggpairs coloured by period
ggpairs(df_gpairs,
        columns = 1:3,
        aes(color = Period, alpha = 0.5)) +
  labs(title = "Pairwise Scatterplots: Demand, Solar Generation , Temperature") +
  theme_minimal()

help(ggpairs)
help(aes)

# 3. ggpairs with log(solar_generation) by period
ggpairs(df_gpairs,
        columns = c(1, 4, 3),   # demand, log(solar), min_temp
        aes(color = Period, alpha = 0.5)) +
  labs(title = "Pairwise Scatterplots: Demand, Log(Solar Generation), Temperature") +
  theme_minimal()
```
# Final Multivariate Plots (Josh)

## Pairwise Correlation
```{r}
# # data
# str(df_ext)
# str(df_gpairs)
# 
# # Include groupings by weekday and weekend
# df_gpairs <- df_gpairs %>%
#   mutate(
#     weekday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
#   )


# 3. ggpairs with log(solar_generation) by period
# demand solar wind and temp
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships

)

# demand solar wind and temp
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "solar_generation", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman")) # spearman allows for non-linear relationships (p value is approx)

)
```
```{r}
# Create a new variable "day_type" for weekend vs weekday
df_gpairs$date <- df_ext$date

df_gpairs <- df_ext %>%
  mutate(
    day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )

str(df_gpairs)
str(df_ext)

# 1a: Combine the grouping of Period and day_type using interaction
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "min_temp"),
  mapping = aes(color = interaction(Period, day_type), alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman"))
)

# 1b) Use solar_generation instead of log_solar_gen
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "solar_generation", "min_temp"),
  mapping = aes(color = interaction(Period, day_type), alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman"))
)

# 2A) No weekend/weekday. color only by Period, using log_solar_gen
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "log_solar_gen", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman"))
)

# 2B) No weekend/weekday. color only by Period, using solar_generation
ggpairs(
  data = df_gpairs,
  columns = c("national_demand", "solar_generation", "min_temp"),
  mapping = aes(color = Period, alpha = 0.5),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "points"),
  upper = list(continuous = wrap("cor", method = "spearman"))
)



```

## Violin 
```{r}
df_var1 <- data.frame(
  var = "national_demand",
  val = df_ext$national_demand
)
df_var2 <- data.frame(
  var = "solar_generation",
  val = df_ext$solar_generation
)

#  Set axis 
demandMin <- 22500
demandMax <- 45000  

solarMin  <- 0
solarMax  <- 1600   

# Compute scaleFactor so [solarMin, solarMax] -> [demandMin, demandMax]
scaleFactor <- (demandMax - demandMin) / (solarMax - solarMin)

#  Build the plot, manually transforming solar
ggplot() +
  #National Demand Violin 
  geom_violin(
    data = df_var1,
    aes(x = var, y = val),
    fill = "grey80", alpha = 0.5, color = "black", linewidth = 0.5
  ) +
  
  # Solar Generation Violin ----
  geom_violin(
    data = df_var2,
    # transform solar from [solarMin, solarMax] into [demandMin, demandMax]
    aes(x = var, y = (val - solarMin) * scaleFactor + demandMin),
    fill = "red", alpha = .9, color = "black", linewidth = 0.5
  ) +
  
  #  X scale with two categories ===
  scale_x_discrete(limits = c("national_demand", "solar_generation"), name = NULL) +
  
  # === 5) Y scale: left axis [demandMin, demandMax], right axis [solarMin, solarMax] ===
  scale_y_continuous(
    name = "national_demand",
    limits = c(demandMin, demandMax),
    
    sec.axis = sec_axis(
      # invert the scaling for axis labels
      trans = ~ (. - demandMin) / scaleFactor + solarMin,
      name = "solar_generation"
    )
  ) +
  
  # === 6) Style the right axis in red, remove or adjust other theme elements as you like
  theme_minimal() +
  theme(
    axis.text.y.right  = element_text(color = "red"),
    axis.title.y.right = element_text(color = "red"),
    axis.line.y.right  = element_line(color = "red"),
    axis.ticks.y.right = element_line(color = "red")
  )
```

## Violin and Box

```{r}

## Violin and Box

library(ggtext)

# Create datasets for full period and seasonal splits
df_varAll_demand <- data.frame(
  var = "All_demand",
  val = df_ext$national_demand
)

df_varAll_solar <- data.frame(
  var = "All_solar",
  val = df_ext$solar_generation
)

df_varApr_demand <- data.frame(
  var = "Apr-Sep_demand",
  val = df_ext %>% filter(Period == "Apr–Sep") %>% pull(national_demand)
)

df_varApr_solar <- data.frame(
  var = "Apr-Sep_solar",
  val = df_ext %>% filter(Period == "Apr–Sep") %>% pull(solar_generation)
)

df_varOct_demand <- data.frame(
  var = "Oct-Mar_demand",
  val = df_ext %>% filter(Period == "Oct–Mar") %>% pull(national_demand)
)

df_varOct_solar <- data.frame(
  var = "Oct-Mar_solar",
  val = df_ext %>% filter(Period == "Oct–Mar") %>% pull(solar_generation)
)

# Combine into one table
df_all <- dplyr::bind_rows(
  df_varAll_demand,
  df_varAll_solar,
  df_varApr_demand,
  df_varApr_solar,
  df_varOct_demand,
  df_varOct_solar
)

# Set axis ranges and scale factor for solar conversion
demandMin  <- 22500
demandMax  <- 45000
solarMin   <- 0
solarMax   <- 1600
scaleFactor <- (demandMax - demandMin) / (solarMax - solarMin)

# Plot
ggplot() +
  
  # Violin for full-period demand
  geom_violin(
    data = df_all %>% filter(var == "All_demand"),
    aes(x = var, y = val),
    fill = "grey80",
    alpha = 0.5,
    color = "black",
    linewidth = 0.5
  ) +
  
  # Boxplots for seasonal demand
  geom_boxplot(
    data = df_all %>% filter(var %in% c("Apr-Sep_demand", "Oct-Mar_demand")),
    aes(x = var, y = val),
    width = 0.4,
    outlier.shape = 16
  ) +
  
  # Violin for full-period solar (scaled)
  geom_violin(
    data = df_all %>% filter(var == "All_solar"),
    aes(x = var, y = (val - solarMin) * scaleFactor + demandMin),
    fill = "red",
    alpha = 1,
    color = "black",
    linewidth = 0.5
  ) +
  
  # Boxplots for seasonal solar (scaled)
  geom_boxplot(
    data = df_all %>% filter(var %in% c("Apr-Sep_solar", "Oct-Mar_solar")),
    aes(x = var, y = (val - solarMin) * scaleFactor + demandMin),
    width = 0.4,
    outlier.shape = 16,
    color = "red"
  ) +
  
  # X-axis labels
  scale_x_discrete(
    name = NULL,
    limits = c(
      "All_demand",
      "Apr-Sep_demand",
      "Oct-Mar_demand",
      "All_solar",
      "Apr-Sep_solar",
      "Oct-Mar_solar"
    )
  ) +
  
  # Y-axis scales (primary: demand, secondary: solar)
  scale_y_continuous(
    name = "national_demand",
    limits = c(demandMin, demandMax),
    sec.axis = sec_axis(
      ~ (. - demandMin) / scaleFactor + solarMin,
      name = "solar_generation"
    )
  ) +
  
  # Theme adjustments
  theme_minimal() +
  theme(
    axis.text.y.right  = element_text(color = "red"),
    axis.title.y.right = element_text(color = "red")
  )





```

## Time Series Multivariate (
```{r}
# Create smoothed versions of demand and solar
df_smoothed <- df_ext %>%
  mutate(
    demand_smooth = rollmean(national_demand, k = 90, fill = NA, align = "right"),
    solar_smooth  = rollmean(solar_generation,  k = 90, fill = NA, align = "right")
  )

demandCol <- "black"
solarCol  <- "#D55E00"

# Compute scale factor for second axis
demandMin <- min(df_smoothed$demand_smooth, na.rm = TRUE)
demandMax <- max(df_smoothed$demand_smooth, na.rm = TRUE)
solarMin  <- min(df_smoothed$solar_smooth,  na.rm = TRUE)
solarMax  <- max(df_smoothed$solar_smooth,  na.rm = TRUE)

sf <- (demandMax - demandMin) / (solarMax - solarMin)

ggplot() +
  # Demand line
  geom_line(
    data = df_smoothed,
    aes(x = date, y = demand_smooth),
    color = demandCol,
    size  = 1
  ) +
  # Solar line (scaled to demand range)
  geom_line(
    data = df_smoothed,
    aes(x = date, y = (solar_smooth - solarMin) * sf + demandMin),
    color = solarCol,
    size  = 1
  ) +
  labs(
    # title = "Time Series: Demand & Solar Generation (smoothed)",
    x = "Date", 
    y = "National Demand (MW)"
  ) +
  scale_y_continuous(
    # Left axis for Demand
    limits = c(demandMin, demandMax),
    # Right axis for Solar
    sec.axis = sec_axis(
      trans = ~ (. - demandMin) / sf + solarMin,
      name  = "Solar Gen (MW)"
    )
  ) +
  theme_minimal(base_size = 13) +
  # Color-match the axes with the lines
  theme(
    axis.title.y  = element_text(color = demandCol),
    axis.text.y   = element_text(color = demandCol),
    axis.title.y.right = element_text(color = solarCol),
    axis.text.y.right  = element_text(color = solarCol)
  )

```

## Scatter Plots + Linear Fit
```{r}
## 2) Three scatter plots with geom_smooth(method = "lm")

# (A) National Demand (y) vs. Solar Generation (x)
ggplot(df_ext, aes(x = solar_generation, y = national_demand)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Demand vs. Solar Generation", x = "Solar Generation", y = "National Demand") +
  theme_minimal()

# (B) National Demand (y) vs. min_temp (x)
ggplot(df_ext, aes(x = min_temp, y = national_demand)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Demand vs. Min Temperature", x = "Min Temp", y = "National Demand") +
  theme_minimal()

# (C) Solar Generation (y) vs. min_temp (x)
ggplot(df_ext, aes(x = min_temp, y = solar_generation)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Solar Generation vs. Min Temperature", x = "Min Temp", y = "Solar Generation") +
  theme_minimal()

```

## Scatter Plots + Linear Fit + Period
```{r}
# 3) Three scatter plots + lm lines, color by Period

# 1A) Demand vs. Solar Gen, single x-axis colored by Period
ggplot(df_ext, aes(x = solar_generation, y = national_demand, color = Period)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Solar Generation(MV)", y = "National Demand(MV)") +
  theme_minimal()
df_ext

# 1A) Demand vs. Solar Gen, single x-axis colored by Period
dem_sol <- ggplot(df_ext, aes(x = solar_generation, y = national_demand, color = Period)) +
  geom_point(alpha = 0.5) +
  labs(x = "Solar Generation(MV)", y = "National Demand (MV)") +
  # Loess for Apr–Sep only
  geom_smooth(
    data = subset(df_ext, Period == "Apr–Sep"),
    method = "lm", 
    se = TRUE
  ) +
  theme_minimal()
ggMarginal(
  dem_sol,
  type = "boxplot",
  margins = "x",           # Only on the top/bottom margin
  groupColour = TRUE,
  groupFill   = TRUE
)

# Check the summary
mod1 <- lm(national_demand ~ solar_generation, data = df_ext)
summary(mod1)

# Apr–Sep only
mod_aprsep <- lm(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Apr–Sep"))
summary(mod_aprsep)

# Oct–Mar only
mod_octmar <- lm(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Oct–Mar"))
summary(mod_octmar)

# log version
ggplot(df_ext, aes(x = log_solar_gen, y = log_national_demand, color = Period)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  labs(x = "Log Solar Generation", y = "Log National Demand") +
  theme_minimal()
help(geom_smooth)
help(log1p)

# 1B) Demand vs. MinTemp, single x-axis colored by Period
dem_temp <- ggplot(df_ext, aes(x = min_temp, y = national_demand, color = Period)) +
  geom_point(alpha = 0.5) +
  # Loess for Apr–Sep only
  geom_smooth(
    data = subset(df_ext, Period == "Apr–Sep"),
    method = "loess", 
    se = TRUE
  ) +
  # LM for Oct–Mar only
  geom_smooth(
    data = subset(df_ext, Period == "Oct–Mar"),
    method = "lm", 
    se = TRUE
  ) +
  labs(x = "Min Temp(°C)", y = "National Demand (MV)") +
  theme_minimal()
ggMarginal(dem_temp, type = "violin", groupColour = TRUE, groupFill = TRUE)

help("lm")
help("loess")
# Check significance of quadratic term
# Subset data for Apr–Sep
df_aprsep <- subset(df_ext, Period == "Apr–Sep")

# Fit a model with linear + quadratic terms
mod_quad <- lm(national_demand ~ min_temp + I(min_temp^2), data = df_aprsep)

# Look at coefficient estimates and significance
summary(mod_quad)


#Check the summary
mod2_lm <- lm(national_demand ~ min_temp, data = df_ext)
summary(mod2_lm)

mod2_loess <- loess(national_demand ~ min_temp, data = df_ext)
summary(mod2_loess)

# Apr–Sep only
mod_aprsep <- lm(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Apr–Sep"))
summary(mod_aprsep)

# Apr–Sep only
mod_aprsep_loes <- loess(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Apr–Sep"))
summary(mod_aprsep_loes)

# Oct–Mar only
mod_octmar <- loess(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Oct–Mar"))
summary(mod_octmar)

# 1B2 Log version for lcheck
dem_temp <- ggplot(df_ext, aes(x = log1p(min_temp), y = national_demand, color = Period)) +
  geom_point(alpha = 0.5) +
  # Loess for Apr–Sep only
  geom_smooth(
    data = subset(df_ext, Period == "Apr–Sep"),
    method = "l", 
    se = TRUE
  ) +
  # LM for Oct–Mar only
  geom_smooth(
    data = subset(df_ext, Period == "Oct–Mar"),
    method = "lm", 
    se = TRUE
  ) +
  labs(x = "Min Temp(°C)", y = "National Demand (MV)") +
  theme_minimal()
ggMarginal(dem_temp, type = "violin", groupColour = TRUE, groupFill = TRUE)

help("lm")
help("loess")
# Check significance of quadratic term
# Subset data for Apr–Sep
df_aprsep <- subset(df_ext, Period == "Apr–Sep")

# Fit a model with linear + quadratic terms
mod_quad <- lm(national_demand ~ min_temp + I(min_temp^2), data = df_aprsep)

# Look at coefficient estimates and significance
summary(mod_quad)


#Check the summary
mod2_lm <- lm(national_demand ~ min_temp, data = df_ext)
summary(mod2_lm)

mod2_loess <- loess(national_demand ~ min_temp, data = df_ext)
summary(mod2_loess)

# Apr–Sep only
mod_aprsep <- lm(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Apr–Sep"))
summary(mod_aprsep)

# Apr–Sep only
mod_aprsep_loes <- loess(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Apr–Sep"))
summary(mod_aprsep_loes)

# Oct–Mar only
mod_octmar <- loess(national_demand ~ solar_generation, 
                 data = subset(df_ext, Period == "Oct–Mar"))
summary(mod_octmar)
```

## Scatter Plots + Linear Fit + Weekend/day
```{r}
## 4) Three scatter plots + lm lines, facet or color by day_type

# (A) Demand vs. Solar Gen, split by day_type
ggplot(df_gpairs, aes(x = solar_generation, y = national_demand, color = day_type)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ day_type) +
  labs(title = "Demand vs. Solar Generation by Day Type", x = "Solar Generation", y = "National Demand") +
  theme_minimal()
help("lm")
stats:
# (B) Demand vs. min_temp, split by day_type
ggplot(df_gpairs, aes(x = min_temp, y = national_demand, color = day_type)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ day_type) +
  labs(title = "Demand vs. Min Temp by Day Type", x = "Min Temp", y = "National Demand") +
  theme_minimal()

# (C) Solar Gen vs. min_temp, split by day_type
ggplot(df_gpairs, aes(x = min_temp, y = solar_generation, color = day_type)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~ day_type) +
  labs(title = "Solar Generation vs. Min Temp by Day Type", x = "Min Temp", y = "Solar Generation") +
  theme_minimal()


```

# Granger and CCF
```{r}
# Select only numeric variables for correlation and multivariate analysis
numeric_vars <- df[, !("date"), with = FALSE]

energy_ts <- ts(df$national_demand, frequency = 365)
solar_ts <- ts(df$solar_generation, frequency = 365)
wind_ts <- ts(df$wind_generation, frequency = 365)
price_ts <- ts(df$average_price_daily, frequency = 365)
wind_speed_ts <- ts(df$wind_speed, frequency = 365)
rain_ts <- ts(df$rain_mm, frequency = 365)
temp_ts <- ts(df$min_temp, frequency = 365)

# =====================================
# 5.2 GRANGER CAUSALITY ANALYSIS

optimal_lag <- VARselect(cbind(energy_ts, solar_ts), lag.max = 150, type = "const")
print(optimal_lag)


grangertest(energy_ts ~ temp_ts, order = 9)

grangertest(solar_ts ~ energy_ts, order = 9)


grangertest(temp_ts ~ solar_ts, order = 9)

grangertest(energy_ts ~ solar_ts, order = 9)


help(grangertest)

# Custom correlation panel (upper panels)
custom_cor <- function(data, mapping, ...) {
  x <- eval_data_col(data, mapping$x)
  y <- eval_data_col(data, mapping$y)
  corr <- cor(x, y, use = "complete.obs")
  corr_text <- sprintf("%.2f", corr)
  
  df <- data.frame(x = 1, y = 1, corr = corr)
  
  ggplot(df, aes(x, y, fill = corr)) +
    geom_tile(color = "white") +
    geom_text(aes(label = corr_text), size = 5) +
    scale_fill_gradient2(
      low = "blue",     # Negative correlation
      mid = "white",    # Zero correlation
      high = "red",     # Positive correlation
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_void() +
    theme(legend.position = "right")
}


my_column_labels <- c(
  "National\nDemand",
  "Wind\nGen",
  "Solar\nGen",
  "Min\nTemp",
  "Max\nTemp",
  "Rain\nmm",
  "Wind\nSpeed",
  "Avg\nPrice\nDaily"
)

options(scipen = 999)

# Create the ggpairs plot
p <- ggpairs(
  numeric_vars,
  columnLabels = my_column_labels,  
  lower = list(
    continuous = wrap("points", alpha = 0.3, size = 0.8, color = "gray40")
  ),
  diag = list(
    continuous = wrap("densityDiag", fill = "gray40", color = "gray40", alpha = 0.3)
  ),
  upper = list(
    continuous = custom_cor
  ),
  title = "Pairwise Scatterplot with Correlation Matrix"
)

# Adjust theme to fix label overlap & give more space
p <- p + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
  axis.text.y = element_text(size = 8),
  strip.text  = element_text(size = 8),
  plot.margin = margin(10, 10, 10, 10)
)


# print(p)


# Set up a grid layout: 3 rows and 2 columns
par(mfrow = c(3, 2))

# Cross-Correlation: Energy Demand vs. Solar Generation
ccf(solar_ts, energy_ts, lag.max = 30, main = "CCF: National Demand vs. Solar Generation")

# Cross-Correlation: Energy Demand vs. Temperature
ccf(energy_ts, temp_ts, lag.max = 365)

# Cross-Correlation: Energy Demand vs. Electricity Price
ccf(energy_ts, price_ts, lag.max = 30, main = "CCF: Energy Demand vs. Electricity Price")


ccf(energy_ts, solar_ts, lag.max = 365, main = "")


# Define lag for seasonal differencing 
seasonal_lag <- 365

# Apply seasonal differencing and properly handle NA values
df$National_Demand_seasdiff <- c(rep(NA, seasonal_lag), diff(df$national_demand, lag=seasonal_lag))
df$Solar_Generation_seasdiff <- c(rep(NA, seasonal_lag), diff(df$solar_generation, lag=seasonal_lag))

df1 <- data.frame(Time = 1:length(df$National_Demand_seasdiff), 
                 National_Demand_seasdiff = df$National_Demand_seasdiff)

data_clean <- na.omit(df)

kpss.test(data_clean$National_Demand_seasdiff)
kpss.test(data_clean$Solar_Generation_seasdiff)

# Apply seasonal differencing and properly handle NA values
data_clean$National_Demand_seasdiff2 <- c(rep(NA, 7), diff(data_clean$National_Demand_seasdiff, lag = 7))

# Remove NA values introduced by differencing
data_clean <- na.omit(data_clean)

kpss.test(data_clean$National_Demand_seasdiff2)

ccf(data_clean$National_Demand_seasdiff2, data_clean$Solar_Generation_seasdiff, lag.max = 365, main = "CCF: Twice diff Energy Demand vs. Diff. Solar Generation (1 Lag = 365 days)")

```

