---
title: "Group1_Assessment2"
author: "06014041"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Dependencies

```{r dependencies}
library(conflicted)  
library(tidyverse)   
library(lubridate)   
library(tibble)     
library(tidyr)   
library(ggplot2)
library(tidyverse)  
library(data.table) 
library(lubridate)  
library(visdat) 
library(naniar)     
library(zoo)        
library(GGally)       
library(corrplot)     
library(ggplot2)      
library(ggcorrplot)   
library(car)          
library(tseries)      
library(forecast) 
library(patchwork)
library(plotly)

# Resolve conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

```

## Introduction
This notebook reproduces our exploratory data analyses of UK daily energy demand (2020–2024) and its relationship to renewable energy sources, weather, and price variables.


## Load data and check quality
```{r}
file_path <- "energy_demand_uk.csv"  
df <- fread(file_path)  
df$V1 <- NULL
# Preview dataset 
str(df)
glimpse(df)
summary(df)

# Convert date column to proper Date format
df$date <- as.Date(df$date, format = "%Y-%m-%d")

# Convert all time series columns to numeric
numeric_cols <- c("national_demand", "wind_generation", "solar_generation", 
                  "min_temp", "max_temp", "rain_mm", "wind_speed", "average_price_daily")
df[, (numeric_cols) := lapply(.SD, as.numeric), .SDcols = numeric_cols]

sapply(df, class)  # Verify data types

#
# 1.2 Handling Missing Data
#

# Count missing values per column
colSums(is.na(df))

# Handle missing data
df[, min_temp := na.approx(min_temp, na.rm = FALSE)]
df[, max_temp := na.approx(max_temp, na.rm = FALSE)]
df[, rain_mm := rollapply(rain_mm, width = 7, FUN = mean, fill = "extend", align = "right")]
df[, wind_speed := na.approx(wind_speed, na.rm = FALSE)]


# Check if missing values arefilled
colSums(is.na(df))

#
# 1.3 Check for Duplicates & Anomalies
#

# Remove duplicate rows if they exist
df <- unique(df)

# Apply log transformation to all numeric variables
df_log <- df %>%
  mutate(across(all_of(numeric_cols), ~log1p(.)))  

# Boxplot after log transformation
df_log %>% 
  pivot_longer(cols = numeric_cols, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(outlier.colour = "red") +
  theme_minimal() +
  coord_flip() +
  ggtitle("Boxplots (Log-Transformed)")


# Identify extreme values using Z-score method
outlier_threshold <- 3  # Standard Z-score threshold
z_scores <- df[, lapply(.SD, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)), .SDcols = numeric_cols]
outliers <- df[rowSums(abs(z_scores) > outlier_threshold) > 0]

# Print num of detected outliers
print(nrow(outliers))

# Manually Winsorising numeric columns (capping extreme values at 1st and 99th percentile)
df[, (numeric_cols) := lapply(.SD, function(x) {
  q_low <- quantile(x, 0.01, na.rm = TRUE)   # 1st percentile
  q_high <- quantile(x, 0.99, na.rm = TRUE)  # 99th percentile
  x <- pmax(q_low, pmin(x, q_high))  # Cap values within this range
  return(x)
}), .SDcols = numeric_cols]file_path <- "energy_demand_uk.csv"  
df <- fread(file_path)  
df$V1 <- NULL
# Preview dataset 
str(df)
glimpse(df)
summary(df)

# Convert date column to proper Date format
df$date <- as.Date(df$date, format = "%Y-%m-%d")

# Convert all time series columns to numeric
numeric_cols <- c("national_demand", "wind_generation", "solar_generation", 
                  "min_temp", "max_temp", "rain_mm", "wind_speed", "average_price_daily")
df[, (numeric_cols) := lapply(.SD, as.numeric), .SDcols = numeric_cols]

sapply(df, class)  # Verify data types

#
# 1.2 Handling Missing Data
#

# Count missing values per column
colSums(is.na(df))

# Handle missing data
df[, min_temp := na.approx(min_temp, na.rm = FALSE)]
df[, max_temp := na.approx(max_temp, na.rm = FALSE)]
df[, rain_mm := rollapply(rain_mm, width = 7, FUN = mean, fill = "extend", align = "right")]
df[, wind_speed := na.approx(wind_speed, na.rm = FALSE)]


# Check if missing values arefilled
colSums(is.na(df))

#
# 1.3 Check for Duplicates & Anomalies
#

# Remove duplicate rows if they exist
df <- unique(df)

# Apply log transformation to all numeric variables
df_log <- df %>%
  mutate(across(all_of(numeric_cols), ~log1p(.)))  

# Boxplot after log transformation
df_log %>% 
  pivot_longer(cols = numeric_cols, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot(outlier.colour = "red") +
  theme_minimal() +
  coord_flip() +
  ggtitle("Boxplots (Log-Transformed)")


# Identify extreme values using Z-score method
outlier_threshold <- 3  # Standard Z-score threshold
z_scores <- df[, lapply(.SD, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)), .SDcols = numeric_cols]
outliers <- df[rowSums(abs(z_scores) > outlier_threshold) > 0]

# Print num of detected outliers
print(nrow(outliers))

# Manually Winsorising numeric columns (capping extreme values at 1st and 99th percentile)
df[, (numeric_cols) := lapply(.SD, function(x) {
  q_low <- quantile(x, 0.01, na.rm = TRUE)   # 1st percentile
  q_high <- quantile(x, 0.99, na.rm = TRUE)  # 99th percentile
  x <- pmax(q_low, pmin(x, q_high))  # Cap values within this range
  return(x)
}), .SDcols = numeric_cols]

summary(df)
str(df)

```

### Comments on single missing values
It is likely that the missing observation was overlooked as 2024 is a leap year, the missing date is 29/02/2024. **Missingness Mechanism to be determined**, we will remove for now.
```{r}
# Remove all rows that contain any NA values
energy_data_clean <- energy_data %>%
  filter(complete.cases(.))

# Check if any missing values remain
colSums(is.na(energy_data_clean))  # Should return 0 for all columns
```


## Univariate EDA
```{r}
# Pivot the data to long format
energy_data_long <- energy_data_clean %>%
  pivot_longer(
    cols = c(national_demand, wind_generation, solar_generation,
             min_temp, max_temp, rain_mm, wind_speed, average_price_daily),
    names_to = "variable",
    values_to = "value"
  )

# Plot each variable as a time series, using facets
ggplot(energy_data_long, aes(x = date, y = value)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Time Series Plots of All Variables",
    x = "Date",
    y = "Value"
  ) +
  theme_bw()
```

## Repeated univariate visualisations (function)
```{r repeated_univariate_functions}
# Colour-blind friendly palette
cbPalette <- c("#000000", '#E69F00', '#56B4E9', '#009E73', 
               '#F0E442', '#0072B2', '#D55E00', '#CC79A7')

# Function to create overlaid histograms
plot_two_hists <- function(df, var1, var2, binwidth = 1) {
  library(ggplot2)
  
  ggplot(df, aes(x = .data[[var1]])) +
    # First histogram
    geom_histogram(
      aes(y = after_stat(density)),
      binwidth = binwidth,
      fill = cbPalette[1],
      alpha = 0.4,
      colour = "black"
    ) +
    # Second histogram
    geom_histogram(
      aes(x = .data[[var2]], y = after_stat(density)),
      binwidth = binwidth,
      fill = cbPalette[2],
      alpha = 0.4,
      colour = "black"
    ) +
    labs(
      x = "Value",
      y = "Density",
      title = paste("Overlaid Histograms of", var1, "and", var2)
    ) +
    theme_minimal()
}

# Function to plot two densities
plot_two_densities <- function(df, var1, var2) {
  library(ggplot2)
  
  ggplot(df, aes(x = .data[[var1]])) +
    # First density
    geom_density(
      fill = cbPalette[1],
      colour = cbPalette[1],
      alpha = 0.4
    ) +
    # Second density
    geom_density(
      aes(x = .data[[var2]]),
      fill = cbPalette[2],
      colour = cbPalette[2],
      alpha = 0.4
    ) +
    labs(
      x = "Value",
      y = "Density",
      title = paste("Overlaid Densities of", var1, "and", var2)
    ) +
    theme_minimal()
}
```



```{r}
energy_data_clean
```


